\section{Experiments}
\label{sec:experiments}

\subsection{Data}
\label{subsec:data}

\paragraph{Training.}
We use the EF-Cambridge Open Language Database
\citep{geertzen2014efcamdat}, a large-scale corpus of learner English
essays annotated with CEFR proficiency (A1--C2) and L1 metadata.
Following the CMCL companion paper \citep{stearns2026artificial}, we
work with the sentence-level release. The splits live at
\texttt{norm-EFCAMDAT-train.csv} and \texttt{norm-EFCAMDAT-remainder.csv}
(together, the training pool) and \texttt{norm-EFCAMDAT-test.csv} (the
held-out set).

\paragraph{Metadata availability.}
EFCAMDAT exposes the learner's first language (\texttt{l1}) and CEFR
level (\texttt{cefr\_level}) at row level. The three transfer corpora
use the same two fields --- \texttt{l1} plus \texttt{cefr\_level}
(\texttt{cefr\_label} for andrew100k). We keep only these two
pieces of metadata as gate input; richer features (WALS/Grambank
typology, morphological tags) are deferred to v07.

The EFCAMDAT L1 inventory used for the gate embedding is the 10 L1s
observed in the training split: Arabic, French, German, Italian,
Japanese, Mandarin, Portuguese, Russian, Spanish, Turkish. Transfer
rows with out-of-inventory L1 route to a learned \texttt{unk} slot.

\paragraph{Zero-shot transfer.}
We evaluate, without further training, on three external learner
corpora released with the CEFR-classification splits used in our wider
work:
\begin{itemize}
  \item \texttt{norm-andrew100k-remainder.csv} --- the Caines 100k
    collection, with per-row CEFR and L1; the L1 inventory overlaps
    strongly with EFCAMDAT (Portuguese, Mandarin, Russian, Spanish,
    Arabic, German, Italian, French, Japanese, Turkish). This is the
    closest same-target, different-register test.
  \item \texttt{norm-CELVA-SP.csv} --- predominantly French-L1 learners
    of English; French is in the EFCAMDAT L1 inventory, so the gate's
    L1 embedding is reusable at test time.
  \item \texttt{norm-KUPA-KEYS.csv} --- a mixed-L1 corpus whose
    inventory (Polish, Portuguese, Spanish, Italian, Greek, Hungarian,
    German, Dutch, Chinese, Bengali, Slovene) only partially overlaps
    with EFCAMDAT. Several L1s (Polish, Greek, Hungarian, Dutch,
    Bengali, Slovene) are out-of-inventory and route to \texttt{unk}.
    This is the hardest test of L1 generalisation and the main
    motivation for v07 (typological-feature gating).
\end{itemize}

\subsection{Models and baselines}
\label{subsec:models}

We compare four configurations at matched data exposure:
\begin{description}
  \item[B0. Native GPT-2 Base.] No further training; pure OpenAI
    release \citep{radford2019gpt2}. The native baseline.
  \item[B1. Learner GPT-2.] The CMCL companion's continually-pre-trained
    GPT-2 on EFCAMDAT with no metadata signal
    \citep{stearns2026artificial}. The data-only learner baseline.
  \item[B2. Learner GPT-2 + metadata concat.] Same continual
    pre-training as B1, but with CEFR and L1 tokens prepended to each
    training sequence. The prompt-level metadata baseline.
  \item[G1 (ours).] Learner GPT-2 with the metadata-aware SDPA-output
    gate of Eq.~\eqref{eq:metadata-gate}. Head-specific, elementwise,
    sigmoid, multiplicative, initialised to pass-through.
\end{description}

B0--B2 together isolate our two claims: that gating adds something
beyond data (B1) and that an architectural gate beats naive metadata
injection (B2). All models use GPT-2 Base with identical tokenisation
and context length (1024). Continual pre-training uses the same
hyperparameters as \citet{stearns2026artificial} --- AdamW, LR 5e-5 with
cosine decay, batch size 16, 3 epochs over the EFCAMDAT pool --- so the
only varying factor is the gate.

\subsection{Metrics}
\label{subsec:metrics}

\paragraph{Stratified perplexity.}
Primary metric. We report token-level perplexity on the EFCAMDAT
held-out set, broken down by (CEFR, L1) cells. For transfer corpora we
report PPL by CEFR bucket where labels exist and aggregate otherwise.
This matches the reframing of perplexity as diagnostic pursued in
\citet{stearns2025nwpaleval}.

\paragraph{Cloze accuracy.}
On a sample of held-out sentences, we mask single content words picked
by POS (noun, verb, adjective, determiner) and score top-1 and top-5
accuracy of predicting the masked word given bilateral context.
Stratified again by (CEFR, L1).

\subsection{Expected compute}
\label{subsec:compute}

B0 requires no training. B1 is the CMCL checkpoint (already trained).
B2 and G1 require one continual-pre-training run each, estimated at
~2 A100-days per run at GPT-2 Base scale. Inference across the three
transfer corpora is negligible. Total budget for the flagship draft:
~5 A100-days. The nine-variant ablation sweep
(\texttt{experiments-ideas/v01--v09}) is an additional ~18 A100-days if
every variant is trained from scratch, or ~6 A100-days if v05--v09 are
trained only under the frozen-backbone regime of v09.
