\section*{Limitations}
\label{sec:limitations}

\paragraph{EFCAMDAT CEFR labels are curriculum-derived.}
The CEFR labels in EFCAMDAT reflect the level at which a text was
written within the EF curriculum, not an independent assessment of the
text itself. This adds noise to the CEFR axis of our gate and limits
the sharpness of claims about proficiency-specific effects. We mitigate
by stratifying along the coarser A1 / A2 / B1 / B2 / C1 buckets that
EFCAMDAT actually exposes (C2 is absent from the test split).

\paragraph{Single base model, single scale.}
The flagship draft uses GPT-2 Base (117M) only. Qwen's evidence for
$G_1$ dominance holds at 1.7B dense and 15B MoE; whether our extension
to metadata-conditioned gating preserves the gain at GPT-2 Base scale
is the first thing to verify. If it does, a scale sweep (Pythia 160M /
410M, then GPT-2 Medium / Large) becomes the obvious follow-up. We
flag this explicitly rather than over-claim.

\paragraph{Only CEFR and L1 as metadata.}
We deliberately restrict the gate input to CEFR proficiency and L1 ---
the two fields available uniformly across all four corpora we use. This
keeps the cross-corpus evaluation clean but leaves out signals that SLA
research treats as informative: L1 typological features
(articles, pro-drop, word order), age of onset, exposure years, and
task conditions. Variant v07 in \texttt{experiments-ideas/} drops in
WALS/Grambank features in place of the L1 ID and is the natural next
step after the flagship.

\paragraph{Missing-metadata behaviour is under-specified.}
At test time, a corpus with no CEFR or L1 labels routes the embedding
to \texttt{unk}, collapsing the gate to a corpus-average. We do not
yet know whether this gracefully degrades to the ungated baseline or
whether the \texttt{unk} slot ends up with a strange learned
distribution that actively hurts inference. This is an empirical
question the flagship run will answer.

\paragraph{The C4200M synthetic-data extension is deferred.}
\texttt{v10-efcamdat-plus-c4200m.md} in the variant catalogue describes
a potentially useful augmentation --- mixing EFCAMDAT with synthetic
corruptions of web text --- but we keep it out of the flagship to stay
focused on the authentic-data story. If zero-shot transfer to
KUPA-KEYS is poor, the C4200M extension becomes a priority.

\section*{Future Work}
\label{sec:future-work}

\paragraph{Error-distribution alignment via GEC + ERRANT.}
The CMCL companion establishes that a data-only learner GPT-2 produces
an error profile \emph{selectively shifted} toward SLA-vulnerable
categories (verbal morphology, SVA, determiners), measured by tagging
automatic corrections with ERRANT
\citep{bryant2019errant}. The natural follow-up to the present
paper is to apply the same pipeline --- generate, correct (e.g.\ with
CoEdIT \citep{raheja2023coedit}), tag with ERRANT, compute
Jensen--Shannon divergence against authentic EFCAMDAT error histograms
--- to our metadata-conditioned gated model. The expected finding is
that G1's per-(CEFR, L1) histogram is closer to the authentic
per-(CEFR, L1) histogram than B1's is to the EFCAMDAT-average
histogram, demonstrating that the gate provides not just lower
perplexity but finer-grained control over the error mix. We keep this
out of the flagship pilot because it adds a heavy pipeline dependency
(CoEdIT model weights, ERRANT parsing), doubles the eval cost, and
ERRANT covers only morphosyntactic errors --- making it a well-scoped
standalone follow-up rather than a core contribution.

\paragraph{Typological-feature gating for unseen L1s.}
Variant v07 replaces the L1 learned embedding with a fixed WALS /
Grambank typological vector, enabling zero-shot inference on
out-of-inventory L1s. This is the critical test for broad applicability
across languages not represented in EFCAMDAT's ten-L1 training set;
partial KUPA-KEYS transfer (which has Polish, Greek, Hungarian, Dutch,
Bengali, and Slovene among its L1s) is the main motivation for
prioritising it after the flagship result is established.

\paragraph{Scaling the gate with the backbone.}
Qwen's evidence for the SDPA-output gate holds at 1.7B--15B. Once the
GPT-2 Base flagship is validated, the obvious progression is Pythia-160M
/ 410M / 1.4B, testing whether metadata-conditioned gating composes
cleanly with increasing backbone capacity.
